|

### 2023-07-17 paper/00-main.Rmd
shitton of work on supplement and figures
more edits... almost done?

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
  # fig.pos = "p",
  # cache = TRUE
)
library(here)
library(tidyverse)
```

>>>>

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
  # fig.pos = "t"
  # cache = TRUE
)
library(here)
library(tidyverse)
```

### 2023-07-17 paper/01-introduction.Rmd
more edits... almost done?

Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) and (2) defensive ends (DEs).
Typically, DTs and DEs are located within the interior of the line and along the edges, respectively; see Figure \ref{fig:formation} (top) for example formation with DTs and DEs.
The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl].
Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive scheme with a NT.
NFL teams often employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) whose primary role is pass coverage.
Note that apart from the formations shown in Figure \ref{fig:formation}, defensive linemen can have the flexibility to line up differently.
For example, a defensive end, depending on their opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage.

```{r formation, fig.width=5, fig.height=4, fig.pos="t", fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom). The dashed line represents the line of scrimmage separating the defense and offense."}
tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),
       y = c(1, 1, 1, 1.01, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
       pos = c("DE", "NT", "DE", "OLB", "ILB", "ILB", "OLB",
               "CB", "S", "S", "CB"),
       d = "3-4 Defense") |> 
  bind_rows(
    tibble(x = c(3:6, c(2.5, 4.5, 6.5), c(2, 4.1, 4.9, 7)),
           y = c(1, 1, 1, 1, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
           pos = c("DE", "DT", "DT", "DE", "OLB", "MLB", "OLB",
                   "CB", "S", "S", "CB"),
           d = "4-3 Defense") 
  ) |> 
  mutate(d = fct_relevel(d, "4-3 Defense", "3-4 Defense")) |> 
  ggplot(aes(x, y)) +
  geom_tile(fill = "white") +
  geom_label(aes(label = pos), size = rel(4)) +
  geom_hline(yintercept = 0.996, linetype = "dashed") +
  facet_wrap(~ d, scales = "free", ncol = 1) +
  coord_cartesian(ylim = c(0.99, 1.03)) +
  theme_light() +
  theme(strip.text = element_text(size = rel(1.2)),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank())
```

In this work, using data made available in the BDB 2023, we present a novel approach to measure the performance of pass rushers. Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb, @reyers2021quarterback], the literature on evaluating pass rushers is scarce. Below, we provide a brief overview of existing pass rush metrics.

>>>>

Our focus of this manuscript is specifically on measuring the performance of defensive linemen in the NFL.
There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) and (2) defensive ends (DEs).
Typically, DTs and DEs are located within the interior of the line and along the edges, respectively; see Figure \ref{fig:formation} (top) for example formation with DTs and DEs.
The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl].
Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage; see Figure \ref{fig:formation} (bottom) for example defensive scheme with a NT.
NFL teams often employ either one NT or two DTs on the interior with DEs along either side of the defensive line.
In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) whose primary role is pass coverage.
Note that apart from the formations shown in Figure \ref{fig:formation}, defensive linemen can have the flexibility to line up differently.
For example, a defensive end, depending on the opposing matchup, may not necessarily be positioned toward the outside of the line of scrimmage.

```{r formation, fig.width=4.8, fig.height=3.5, fig.pos="t", fig.cap="Two common defensive alignments in football: 4-3 defense (top) and 3-4 defense (bottom). The dashed line represents the line of scrimmage separating the defense and offense."}
tibble(x = c(3:5, 3:6 - 0.5, c(2, 3.7, 4.3, 6)),
       y = c(1, 1, 1, 1.01, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
       pos = c("DE", "NT", "DE", "OLB", "ILB", "ILB", "OLB",
               "CB", "S", "S", "CB"),
       d = "3-4 Defense") |> 
  bind_rows(
    tibble(x = c(3:6, c(2.5, 4.5, 6.5), c(2, 4.1, 4.9, 7)),
           y = c(1, 1, 1, 1, 1.01, 1.01, 1.01, 1.02, 1.02, 1.02, 1.02),
           pos = c("DE", "DT", "DT", "DE", "OLB", "MLB", "OLB",
                   "CB", "S", "S", "CB"),
           d = "4-3 Defense") 
  ) |> 
  mutate(d = fct_relevel(d, "4-3 Defense", "3-4 Defense")) |> 
  ggplot(aes(x, y)) +
  geom_tile(fill = "white") +
  geom_label(aes(label = pos), size = rel(4)) +
  geom_hline(yintercept = 0.995, linetype = "dashed") +
  facet_wrap(~ d, scales = "free", ncol = 1) +
  coord_cartesian(ylim = c(0.99, 1.03)) +
  theme_light() +
  theme(strip.text = element_text(size = rel(1.2)),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank())
```

In this work, using data made available in the Big Data Bowl 2023, we present a novel approach to measure the performance of pass rushers. Relative to other aspects of American football, such as QB evaluation [@burke2019deepqb; @reyers2021quarterback], the literature on evaluating pass rushers is scarce. Below, we provide a brief overview of existing pass rush metrics.

### 2023-07-17 paper/03-methods.Rmd
more edits... almost done?

As for the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
These are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
Finally, we control for play-context covariates with respect to the offensive team.
These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line). We account for play-context since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role. For instance, a team may call a short pass that is intended to be thrown early which could limit the potential STRAIN on a play. Or a team may need to throw a deep pass which would require more time and create more STRAIN potential. We do not account for time directly in the model due to the concern that the time it takes for a quarterback to throw the ball is itself a function of both the play call and pressure from pass rushers. Thus, since we do not know the designed play call, we condition on the play context to adjust for play-level differences attributing to a pass rusher's STRAIN.

>>>>

As for the fixed-effects about pass rusher $j$ in play $i$, we include a variety of features that likely contribute to variation in STRAIN.
First, we adjust for the position of both the pass rusher and nearest blocker to account for any positional effects.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
These are encoded as indicator variables with defensive ends and tackles as the reference levels for the pass rushers and blockers, respectively.
We also adjust for the number of pass blockers on the play, since teams may decide to employ a more protective scheme that could lower the observed STRAIN.
Finally, we control for play-context covariates with respect to the offensive team.
These include the current down (first, second, third, fourth, or two-point conversion), yards to go for a first down, and current yardline (i.e., distance from the possession team's goal line). 
We account for play-context since these variables impact a team's designed play, which may result in a play with low or high STRAIN regardless of the pass rusher's role. 
For instance, a team may call a short pass that is intended to be thrown early which could limit the amount of STRAIN on a play. 
Or a team may need to throw a deep pass which would require more time and potentially create more STRAIN.
We do not account for time directly in the model due to the concern that the time it takes for a quarterback to throw the ball is itself a function of both the play call and pressure from pass rushers.
Thus, since we do not know the designed play call, we condition on the play context to adjust for play-level differences attributing to a pass rusher's STRAIN.

### 2023-07-17 paper/04-results.Rmd
more edits... almost done?

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the varying intercept distributions for the top ten pass rushers for each position.
We obtain similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (defensive end), T.J. Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
Additionally, we observe different levels of variability across these top players, with only a subset of players with intercept ranges strictly above zero.
This is not necessarily surprising given the limited sample of data.

>>>>

In addition to adjusting for opposition, our multilevel model allows us to provide some notion of uncertainty about the pass rusher's effect on the play level STRAIN.
Figure \ref{fig:fig_rankings_boot} displays the varying intercept distributions for the top ten pass rushers for each position.
We obtain similar rankings based on the random intercepts in comparison to the average STRAIN results in Tables \ref{tab:edge} and \ref{tab:interior}, with Myles Garrett (defensive end), T.J. Watt (outside linebacker), Aaron Donald (defensive tackle), and Vita Vea (nose tackle) as the first-ranked pass rusher for their respective position.
Additionally, we observe different levels of variability across these top players, with only a subset of players having intercept distributions strictly above zero.
This is not necessarily surprising given the limited sample of data.

### 2023-07-17 paper/05-discussion.Rmd
more edits... almost done?

In this work, we have proposed STRAIN---a simple and interpretable statistic for evaluating pass rushers---with higher values corresponding to greater pass rushing ability.
STRAIN is a model-free metric which can easily be integrated into any data pipeline without much computational cost.
Visualizations of STRAIN can be useful and intuitive for coaches and broadcasters in various aspects ranging from gameplan preparation to in-game real-time play analysis.
We demonstrate that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player tracking data can lead to the development of insightful metrics in American football.

Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession and pass rushing teams, and finally the pass rusher's assigned blocker.
This is an intuitive result, consistent with previous literature on pass rushing and pressure.
However, our model is limited in that we only account for players directly involved as the pass rusher or nearest blocker, on top of team-level effects.
The nested structure of our model enforces positive dependence between pass rushers on the same team, which may not be true.
For instance, if one player is known to be an elite pass rusher then the opposing blocking team may focus their blocking efforts on this player. 
This could leave the path open for another player to rush the quarterback with ease, resulting in higher STRAIN values.
As shown in Figure \ref{fig:fig_rankings_boot}, the first- and fourth-ranked defensive ends Myles Garrett and Takkarist McKinley were teammates on the Cleveland Browns during this period of time.
McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an great pass rusher.
In order to capture this type of behavior, we could consider modeling an aggregate STRAIN across pass rushers with a regularized adjusted plus-minus (RAPM) regression approach.
The use of RAPM techniques has been successfully demonstrated in American football by @sabin2021estimating, which accounts for all players on the field with Bayesian hierarchical models.
We leave this type of analysis for future work, which will require careful consideration of available informative priors [@matano2023augmenting].

>>>>

In this work, we have proposed STRAIN---a simple and interpretable statistic for evaluating pass rushers---with higher values corresponding to greater pass rushing ability.
STRAIN is a model-free metric which can easily be integrated into any data pipeline without much computational cost.
Visualizations of STRAIN can be useful and intuitive for coaches and broadcasters in various aspects ranging from gameplan preparation to in-game real-time play analysis.
We demonstrate that a pass rusher's average STRAIN is both stable and more predictive of future pressure events than using the player's previously observed pressure events.
This is analogous to the use of exit velocity in baseball, a predictive measurement that avoids the noisy nature of observed outcomes, emphasizing how the opportunity of working with player tracking data can lead to the development of insightful metrics in American football. 
Through multilevel modeling, we observe that the pass rusher explains more variation in their play-level STRAIN, followed by the possession and pass rushing teams, and finally the pass rusher's assigned blocker.
This is an intuitive result, consistent with previous literature on pass rushing and defensive pressure.

Our multilevel model, however, is subject to several limitations.
We only account for players directly involved as the pass rusher or nearest blocker, on top of team-level effects.
The nested structure of our model enforces positive dependence between pass rushers on the same team, which may not be true.
For instance, if one player is known to be an elite pass rusher then the opposing blocking team may focus their blocking efforts on this player. 
This could leave the path open for another player to rush the quarterback with ease, resulting in higher STRAIN values.
As shown in Figure \ref{fig:fig_rankings_boot}, the first- and fourth-ranked defensive ends Myles Garrett and Takkarist McKinley were teammates on the Cleveland Browns during the first eight weeks of the 2021 NFL season.
McKinley is a surprising name on our list and we suspect that his high rank is mostly due to playing on the same defensive unit with Garrett, who is well-known as an great pass rusher.
In order to capture this type of behavior, we could consider modeling an aggregate STRAIN across pass rushers with a regularized adjusted plus-minus (RAPM) regression approach.
The use of RAPM techniques has been successfully demonstrated in American football by @sabin2021estimating, which accounts for all players on the field with Bayesian hierarchical models.
We leave this type of analysis for future work, which will require careful consideration of available informative priors [@matano2023augmenting].
